Extending EAGGA to Neural Networks

### 2025-03-07 paper/main.tex
more bullets for paper

regularisation
- LM: ridge (L2) + lasso (L1)
  * ridge: overcome multi-collinearity by pulling coeffs close to 0
  * lasso: variable selection by setting coeffs = 0
- NN:
  * L2 regularisation implicitly included in SGD when using weight decay (pull params towards 0) -> does this combat multi-collinearity as well or serve another purpose?
  * L1
    + key claim of DL: feature selection obsolete, network finds important ones itself
    + still, feature usually always has some impact (back this up), no explicit feature selection
    + explicit feature selection simply by not inluding a feature
  * another classical NN regularisation technique: dropout against co-adaptation
  * early stopping

@@

\section{Extension to neural networks}
- why extend it to NNs?
  * NNs notoriously uninterpretable due to complex transformation of the feature space
  * due to (lin alg) non-linear activation functions, interactions can be modelled
  * no monotonicity guarantees
- our approach
  * feature sparsity: straightforward, only train model on selected features, goes somewhat against idea of deep learning,
    where the model is supposed to decide on its own, which features to put importance on -> WHY?
  * feature interaction
    - create non-fully connected network by introducing network-groups based on the equivalence relations introduced in the EAGGA paper
    - show formula why the interacting features need to be grouped together when using ReLU (cf. photos)
  * monotonicity constraints
    - for groups with contraint on monotonicity: clip weights to [0, infty), bias clipping not necessary
    - show formula of how monotonicity is enforced with this
  * changes (-) + similarities (+) to EAGGA paper (~ in-between)
    + 2/3 holdout outer split for 'CV'- vs test-set
    - train + eval each individual via 5-fold CV with 20\% of train set again reserved to determine early stopping -> best to visualise,
      * early stopping necessary (as opposed to XGB) as NN can be trained forever, whereas tree implementations used by EAGGA paper have
        max depth, etc.
      * early stopping using min-epochs 200, patience 100, i.e. train for $\ge$ 200 epochs, always safe model with lowest loss, whenever
        current loss > mean(losses of last 100 (patience) epochs), stop early + return model with lowest loss
    ~ if no early stopping, train each fold (in 5-fold CV) for max 120 secs, do entire EAGGA loop per dataset for max 8 hrs
    ~ training on Sagemaker Notebook instance
      * initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
      * thus decided for more economic + ressourcenschonend t3.medium
    - evaluation
      * AUC + NF: straightforward
      * NI: for each group with $\ge$ 2 features: sum from 1 to (\# features - 1), i.e ${\# features \choose 2}$,
        then sum over all groups like this, then divide by p(p-1)/2, i.e. ${p \choose 2}$
      * NNM: \# unconstrained features / \# all features
    ~ feature + interaction detector: examination of decision trees not super straightforward in sklearn, hence simply use p=0.5
    ? interaction detector (FAST)
      * train logistic model on 80\% of data (only holdout train portion or all ??), evaluate on 20
      * use mean accuracy (logistic metric) instead of RSS (LM metric)
    + monotonicity detector as in paper, also use dec tree HPs from mlr3
    ~ encoding of group structures similar, each feature will get its individual monotonicity sign from the detector, this is then fed to the dataset,
      which then outputs the data according multiplied with the sign

>>>>

regularisation
- LM: ridge (L2) + lasso (L1)
  * ridge: overcome multi-collinearity by pulling coeffs close to 0
  * lasso: variable selection by setting coeffs = 0
- NN:
  * L2 regularisation implicitly included in SGD when using weight decay (pull params towards 0) -> does this combat multi-collinearity as well or serve another purpose?
  * L1
    + key claim of DL: feature selection obsolete, network finds important ones itself
    + still, feature usually always has some impact (back this up), no explicit feature selection
    + explicit feature selection simply by not inluding a feature
  * another classical NN regularisation technique: dropout against co-adaptation
  * early stopping against overfitting when using iterative method

@@

\section{Extension to neural networks}
- why extend it to NNs?
  * NNs notoriously uninterpretable due to complex transformation of the feature space
  * due to (lin alg) non-linear activation functions, interactions can be modelled
  * no monotonicity guarantees
- our approach
  (general algorithm)
  * EAGGA algorithm implemented just as described in the paper
  * group structure init
    - feature detector
      * as described in paper
      * but instead of fitting 10 trees + taking rel. \# of features used as p for trunc geom, we simply use 0.5 (sklearn dectree analysis a bit tricky)
      * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we only use the features with non-0 filter values
    - interaction detector
      * as described in paper
      * same as feat detect, use p=0.5 to sample \# of interactions used instead of 10 dec trees
      * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
    - monotonicity detector
      * as described in paper
      * use default HPs (max depth 30, minsplie = 20) of mlr3 classification tree implementation, as this is the library the paper uses
    -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
  (implementation details)
  * group structures + datasets
  * neural network
    - "instead of XGB we use NN, ..., this requires special architecture, etc. pp."
  * holdout, cv, early stopping

  (old notes, use in new)
  * feature sparsity: straightforward, only train model on selected features, goes somewhat against idea of deep learning,
    where the model is supposed to decide on its own, which features to put importance on -> WHY?
  * feature interaction
    - create non-fully connected network by introducing network-groups based on the equivalence relations introduced in the EAGGA paper
    - show formula why the interacting features need to be grouped together when using ReLU (cf. photos)
  * monotonicity constraints
    - for groups with contraint on monotonicity: clip weights to [0, infty), bias clipping not necessary
    - show formula of how monotonicity is enforced with this
  * changes (-) + similarities (+) to EAGGA paper (~ in-between)
    + 2/3 holdout outer split for 'CV'- vs test-set
    - train + eval each individual via 5-fold CV with 20\% of train set again reserved to determine early stopping -> best to visualise,
      * early stopping necessary (as opposed to XGB) as NN can be trained forever, whereas tree implementations used by EAGGA paper have
        max depth, etc.
      * early stopping using min-epochs 200, patience 100, i.e. train for $\ge$ 200 epochs, always safe model with lowest loss, whenever
        current loss > mean(losses of last 100 (patience) epochs), stop early + return model with lowest loss
    ~ if no early stopping, train each fold (in 5-fold CV) for max 120 secs, do entire EAGGA loop per dataset for max 8 hrs
    ~ training on Sagemaker Notebook instance
      * initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
      * thus decided for more economic + ressourcenschonend t3.medium
    - evaluation
      * AUC + NF: straightforward
      * NI: for each group with $\ge$ 2 features: sum from 1 to (\# features - 1), i.e ${\# features \choose 2}$,
        then sum over all groups like this, then divide by p(p-1)/2, i.e. ${p \choose 2}$
      * NNM: \# unconstrained features / \# all features
    ~ feature + interaction detector: examination of decision trees not super straightforward in sklearn, hence simply use p=0.5
    ? interaction detector (FAST)
      * train logistic model on 80\% of data (only holdout train portion or all ??), evaluate on 20
      * use mean accuracy (logistic metric) instead of RSS (LM metric)
    + monotonicity detector as in paper, also use dec tree HPs from mlr3
    ~ encoding of group structures similar, each feature will get its individual monotonicity sign from the detector, this is then fed to the dataset,
      which then outputs the data according multiplied with the sign
